---
title: "Lab 10: Distances and the Border Zone"
subtitle: "ESS 330"
author:
  - name: "Maya McCain"
    email: "maya.mccain@colostate.edu"
format: html
  self-contained: true 
editor: visual
---

# Libraries
```{r}
# spatial data science
library(tidyverse)
library(sf)
library(units)

# Data
library(AOI)

# Visualization
library(gghighlight)
library(ggrepel)
library(knitr)

library(rnaturalearth)
```

# Question 1
## 1.1 - Define a Projection
```{r}
eqdc <- '+proj=eqdc +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs'
```

## 1.2 - Get USA state boundaries
```{r}
remotes::install_github("mikejohnson51/AOI")
remotes::install_github("ropensci/USAboundaries")
remotes::install_github("ropensci/USAboundariesData")
remotes::install_github("ropenscilabs/rnaturalearthdata")
```
## 1.3 - Get country boundaries for Mexico, the United States of America, and Canada
```{r}
aoi_get(country = c("MX", "CA", "USA"))

eqdc <- st_crs(5070)

countries <- rnaturalearth::countries110 |>
  st_transform(eqdc)

states <- USAboundaries::us_states(resolution = "low") |>
  filter(!state_abbr %in% c("HI", "AK", "PR")) |>
  st_transform(eqdc)
```

## 1.4 - Get city locations from the CSV file
```{r}
library(readr)
cities <- readr::read_csv("data/simplemaps_uscities_basicv1.90/uscities.csv", show_col_types = FALSE) |>
  st_as_sf(coords = c("lng", "lat"), crs = 4326) |>
  filter(!state_id %in% c("HI", "AK", "PR")) |>
  st_transform(eqdc) 
```

# Question 2
## 2.1 - Distance to USA Border (coastline or national) (km)
```{r}
state_union <- st_union(states) %>%
  st_cast("MULTILINESTRING") %>%
  st_as_sf()

border_cities <- cities %>%
  st_filter(st_union(states)) %>%
  filter(!state_id %in% c("HI", "AK", "PR")) %>% 
  mutate(dist_to_border = set_units(st_distance(., state_union), "km"))

select(border_cities, city, state_name, dist_to_border) %>%
  st_drop_geometry() %>%
  slice_max(dist_to_border, n = 5) %>%
  flextable::flextable()

```

## 2.2 - Distance to States (km)
```{r}
state_combine <- st_combine(states) %>%
  st_cast("MULTILINESTRING") %>%
  st_as_sf()

state_cities <- cities %>%
  st_filter(st_combine(states)) %>%
  filter(!state_id %in% c("HI", "AK", "PR")) %>%
  mutate(dist_to_state = set_units(st_distance(., state_combine), "km"))

select(state_cities, city, state_name, dist_to_state) %>%
  st_drop_geometry() %>%
  slice_max(dist_to_state, n = 5) %>%
  flextable::flextable()
```

## 2.3 - Distance to Mexico (km)
```{r}
mexico <- countries %>% 
  filter(NAME == "Mexico")

mexico_combine <- st_combine(mexico) %>%
  st_cast("MULTILINESTRING") %>%
  st_as_sf()

mex_cities <- cities %>%
  st_filter(st_combine(states)) %>%
  filter(!state_id %in% c("HI", "AK", "PR")) %>%
  mutate(dist_to_mexico = set_units(st_distance(., mexico_combine), "km"))

select(mex_cities, city, state_name, dist_to_mexico) %>%
  st_drop_geometry() %>%
  slice_max(dist_to_mexico, n = 5) %>%
  flextable::flextable()
```
## 2.4 - Distance to Canada (km)
```{r}
canada <- countries %>% 
  filter(NAME == "Canada")

canada_combine <- st_combine(canada) %>%
  st_cast("MULTILINESTRING") %>%
  st_as_sf()

canada_cities <- cities %>%
  st_filter(st_combine(states)) %>%
  filter(!state_id %in% c("HI", "AK", "PR")) %>%
  mutate(dist_to_canada = set_units(st_distance(., canada_combine), "km"))

select(canada_cities, city, state_name, dist_to_canada) %>%
  st_drop_geometry() %>%
  slice_max(dist_to_canada, n = 5) %>%
  flextable::flextable()
```
# Question 3
## 3.1 Data
```{r}
n_america <- countries %>%
  filter(NAME %in% c("United States of America", "Canada", "Mexico")) %>%
  st_transform(5070)

conus_states <- AOI::aoi_get(state = "conus") %>%
  st_transform(5070) %>% 
  st_cast("MULTILINESTRING")

conus_outline <- AOI::aoi_get(state = "conus") %>%
  filter(!state_abbr %in% c("HI", "AK", "PR")) %>%
  st_union() %>%
  st_cast("MULTILINESTRING") %>% 
  st_as_sf()

top_10_cities <- cities |> 
  slice_max(population, n = 10) |> 
  st_transform(5070)
```

```{r}
library(ggthemes)
ggplot() + 
  geom_sf(data = n_america, lty = 3, alpha = .5) + 
  geom_sf(data = conus_states, size = 1, fill = "NA", color = "grey") +
  geom_sf(data = conus_outline, size = 1.5, fill = "NA", color = "black") +
  geom_sf(data = top_10_cities, size= 2, color = "red") +
  ggrepel::geom_label_repel(
    data = top_10_cities,
    aes(label = city, geometry = geometry),
    stat = "sf_coordinates",
    size = 3) +
  ggthemes::theme_map() 
```
## 3.2 City Distance from the Border
```{r}
border_5_cities <- border_cities %>% 
  slice_max(dist_to_border, n = 5)
ggplot() + 
  geom_sf(data = n_america, lty = 3, alpha = .5) + 
  geom_sf(data = conus_states, size = 1, fill = "NA", color = "gray") +
  geom_sf(data = conus_outline, size = 1.5, fill = "NA", color = "black") +
  geom_sf(data = border_cities, aes(color = as.numeric(dist_to_border))) +
  geom_sf(data = border_5_cities, size= 2, color = "red") +
  ggrepel::geom_label_repel(
    data = border_5_cities,
    aes(label = city, geometry = geometry),
    stat = "sf_coordinates",
    size = 3) +
  ggthemes::theme_map() 
```
## 3.3 City Distance from Nearest State
```{r}
state_5_cities <- state_cities %>% 
  slice_max(dist_to_state, n = 5)
ggplot() + 
  geom_sf(data = n_america, lty = 3, alpha = .5) + 
  geom_sf(data = conus_states, size = 1, fill = "NA", color = "gray") +
  geom_sf(data = conus_outline, size = 1.5, fill = "NA", color = "black") +
  geom_sf(data = state_cities, aes(color = as.numeric(dist_to_state))) +
  geom_sf(data = state_5_cities, size= 2, color = "red") +
  ggrepel::geom_label_repel(
    data = state_5_cities,
    aes(label = city, geometry = geometry),
    stat = "sf_coordinates",
    size = 3) +
  ggthemes::theme_map() 
```

## 3.4 Equidistance boundary from Mexico and Canada
```{r}
cities <- cities %>%
  mutate(dist_to_mexico = set_units(st_distance(., mexico_combine), "km")) %>% 
  mutate(dist_to_canada = set_units(st_distance(., canada_combine), "km")) %>% 
  mutate(dist_to_border = set_units(st_distance(., state_union), "km"))
```


```{r}
eqdist <- cities %>% 
  mutate(equidist_diff = as.numeric(dist_to_mexico) - as.numeric(dist_to_canada))

eqdist_zone <- eqdist %>%
  mutate(abs_diff = abs(equidist_diff)) %>%
  slice_min(abs_diff, n = 20)

top_5_eqdist <- eqdist_zone %>%
  slice_max(population, n = 5)
```

```{r}
ggplot() + 
  geom_sf(data = n_america, lty = 3, alpha = .5) + 
  geom_sf(data = conus_states, size = 1, fill = "NA", color = "grey") +
  geom_sf(data = conus_outline, size = 1.5, fill = "NA", color = "black") +
  geom_sf(data = eqdist) +
  gghighlight::gghighlight(abs(equidist_diff) < 100) +
  geom_sf(data = top_5_eqdist, size= 2, color = "red") +
  ggrepel::geom_label_repel(
    data = top_5_eqdist,
    aes(label = city, geometry = geometry),
    stat = "sf_coordinates",
    size = 3) +
  ggthemes::theme_map() 
```
# Question 4
## 4.1 Quantifing Border Zone
```{r}
border_zone <- cities %>%
  filter(as.numeric(dist_to_border) <= 160)

n_border_zone <- nrow(border_zone)

border_zone_pop <- border_zone %>%
  st_drop_geometry() %>%
  summarise(total_population = sum(population, na.rm = TRUE)) %>%
  pull(total_population)

total_pop <- cities %>%
  st_drop_geometry() %>%
  summarise(total_population = sum(population, na.rm = TRUE)) %>%
  pull(total_population)

pop_percentage <- (border_zone_pop / total_pop) * 100

summary_df <- data.frame(
  `Number of Cities in Border Zone` = n_border_zone,
  `Population in Border Zone` = border_zone_pop,
  `Total Population (All Cities)` = total_pop,
  `Percent in Border Zone (%)` = round(pop_percentage, 2)
)

flextable::flextable(summary_df)

```
This information matches the ACLU estimate that about two-thirds of the population lives in the danger zone (within 100 miles of the US border). 

## 4.2 Mapping Border Zone
```{r}
border_zone <- cities %>%
  filter(as.numeric(dist_to_border) <= 160)

danger_top10 <- border_zone %>%
  slice_max(population, n = 10)

ggplot() +
  geom_sf(data = cities, aes(color = population), alpha = 0.4, size = 0.8) +
  gghighlight::gghighlight(as.numeric(dist_to_border) <= 160,
                           label_key = city,
                           use_direct_label = FALSE,
                           keep_scales = TRUE) +
  scale_color_gradient(low = "orange", high = "darkred", name = "Distance to Border") +
  geom_sf(data = danger_top10, color = "black", size = 2) +
  ggrepel::geom_label_repel(
    data = danger_top10,
    aes(label = city, geometry = geometry),
    stat = "sf_coordinates",
    size = 3,
    fill = "white"
  ) +
  labs(title = "U.S. Cities Within 100 Miles of the Border") +
  ggthemes::theme_map()

```

## 4.3 : Instead of labeling the 10 most populous cites, label the most populous city in each state within the Danger Zone
```{r}
border_zone <- cities %>%
  filter(as.numeric(dist_to_border) <= 160)

top_city_per_state <- border_zone %>%
  group_by(state_name) %>%
  slice_max(population, n = 1, with_ties = FALSE) %>%
  ungroup()

ggplot() +
  geom_sf(data = border_zone, aes(color = population), alpha = 0.6, size = 1) +
  scale_color_gradient(low = "orange", high = "darkred", name = "Distance to Border") +
  geom_sf(data = top_city_per_state, color = "black", size = 2) +
  ggrepel::geom_label_repel(
    data = top_city_per_state,
    aes(label = city, geometry = geometry),
    stat = "sf_coordinates",
    size = 3,
    fill = "white"
  ) +
  labs(
    title = "Most Populous Cities in Each State Within 100 Miles of the Border"
  ) +
  ggthemes::theme_map()
```

